Copulas and their potential for ecology

### 2019-10-22 Paper.Rmd
ready to submit, I think

---
title: "Copulas and their potential for ecology"
author: "Shyamolina Ghosh, Lawrence W. Sheppard, Mark T. Holder, Terrance E. Loecke, Philip C. Reid, James D. Bever, Daniel C. Reuman"
date: ""
fontsize: 12pt
geometry: "left=2.54cm,right=2.54cm,top=2.54cm,bottom=2.54cm"

output: 
  pdf_document:
    number_sections: no
    keep_tex: yes
    fig_caption: yes
    includes:
      in_header: head_maintext.sty

@@

\noindent \emph{The number of words in the abstract:} 261

@@

<!--***DAN: remianing to do:
DONE 1) Edit abstract
DONE 2) Go through the whole thing one more time
3) send it out to coauthors (done) and maybe get feedback
DONE 3.5) ESP FROM lAWRENCE
DONE 3.6) Incorporate Shya feedback
DONE 4)make sure to say something about tropical systems
DONE 5) maybe write to the editors saying this is coming soon?
DONE 6) In the Discussion, you maybe need to point out not only that we
use only bivariate copulas (and there are multivariate ones), but also
that we use only continuous marginals. Probably an adequate approximation
for our data, but see the other papers I've seen recently which use count data,
and discrete marginals. Approaches such as that will be necessary for some 
datasets.
DONE 7) change the box around the Cedar Creek data in the main fig
DONE 8) say in Discussion that you could have used non-ranks for the nonparam stats - this
is the stuff the Ecol Mongr referee talked about. Maybe this should go in Background instead?
9) maybe go back and read the Ecol Monogr reviews again
DONE 10) try to emphasize early on that the focus is on positively associated variables, so 'lower tail' means the lower tail of both variables and it is not necessary to worry about negative relationships (in practice we would always transform these for ease).   The dfinitions of tail dependence we use cannot usefully be applied to untransformed data in fig 1e,f.
-->

# Abstract
All branches of ecology study relationships among environmental and biological variables. However, 
standard approaches to studying such relationships, based on correlation and regression, 
provide only a small slice of the complex information contained in the relationships. Other 
statistical approaches exist that provide a complete description of relationships between variables, 
based on the concept of the *copula*; they are applied in finance, neuroscience and other fields, 
but rarely in ecology. We here explore the concepts that underpin copulas and examine the potential 
for those concepts to improve our understanding of ecology. We find that informative copula 
structure in dependencies between variables is common across all the environmental, 
species-trait, phenological, population, community, and ecosystem functioning datasets we 
considered. Many datasets exhibited asymmetric tail associations, whereby two variables were more 
strongly related in their left compared to right tails, or *vice versa*. We describe mechanisms by which 
observed copula structure and asymmetric tail associations can arise in ecological data, including a Moran-like 
effect whereby dependence structures between environmental variables are inherited by ecological 
variables; and asymmetric or nonlinear influences of environments on ecological variables, such as 
under Liebig's law of the minimum. We also describe consequences of copula structure for 
ecological phenomena, including impacts 
on extinction risk, Taylor's law, and the stability through time of ecosystem services. By documenting the 
importance of a complete description of dependence between variables, advancing conceptual frameworks, and 
demonstrating a powerful approach, we aim to encourage widespread use of copulas in ecology, which we
believe can benefit the discipline.

@@

All branches of ecology 
study relationships among environmental and biological 
variables. However, commonly used correlation and regression 
approaches to studying 
such relationships are limited, and 
provide only a partial description of a relationship between variables. 
For instance, datasets showing markedly different 
relationships may have the same correlation coefficient
(Fig. \ref{fig_cop_pedag1}A-B). The variables of Fig. \ref{fig_cop_pedag1}A 
(respectively, Fig. \ref{fig_cop_pedag1}B) are principally 
related in the left (respectively, right) portions of their distributions, 
an asymmetric pattern of dependence that can
have ecological significance, as discussed below, but that is not captured by 
correlation. Although correlation is only one way to 
study relationships between variables, other common metrics also 
provide only partial information.

@@

Well-developed approaches do exist, however, and are 
applied widely in other fields, 
that provide a description of the dependence 
between two or more variables that can be mathematically proven 
to be complete [@nelsen2006_copula; @joe2014_dependence; @MaiScherer2017]. 
The approaches are based on 
the concept of the *copula*. Copula approaches separate the information of a bivariate 
random variable into two distinct parts: the information in the 
marginal distributions (which says nothing about the dependence between the 
variables), and the remaining information (which is solely about
dependence). We here introduce some simple concepts, based on taking ranks,
that relate to copulas and give a conceptual 
flavor of copula ideas to be introduced 
more formally in the section on Background below. Given 
a sample $(x_i,y_i)$, $i=1,2,\ldots,n$ (e.g., 
Fig. \ref{fig_cop_pedag1}A-B), information about the structure of the dependence
between $x$ and $y$ can be separated from information contained 
in the marginal distributions by 
considering the bivariate plot of $u_i$ versus $v_i$, where
$u_i$ is the rank of $x_i$ in the set $x_j$ ($j=1,\ldots,n$), divided by
$n+1$; and $v_i$ is defined in the same way but using the $y_i$. The rank
of the smallest element in a set is $1$. The $u_i$ and 
$v_i$ are called *normalized ranks* of the $x_i$ and $y_i$; they relate to 
the empirical cumulative distribution function of the $x_i$ and $y_i$,
respectively. We henceforth abbreviate "cumulative distribution function" as "distribution
function" or "cdf". See, for instance,
Fig. \ref{fig_cop_pedag1}C-D, which show the normalized ranks of 
A-B, respectively. Ranking makes the marginal distributions of 
the component datasets uniform, isolating the 
dependence structure. Dependence structure and marginals can then be studied 
separately. Normalized rank plots such as Fig. \ref{fig_cop_pedag1}C-D
relate to copulas in a way to be described in the section on Background below, 
after a definition of copulas
is provided. Note that Pearson correlation, even though it is the most commonly
used measure of relationships between variables, is modified by monotonic
tranformation of the component variables (Appendix 
\ref{SM-pedagog1_details}, Fig. \ref{SM-fig_transf_pearson_spearman}),
and therefore reflects not only dependence information, but also
information on the marginals [@Genest2007]. 
Spearman and Kendall correlations, being rank-based, 
are not influenced by monotonic transformations of the component variables. 
They provide information solely on the dependence between those variables.

A main benefit of a copula approach is 
that it can detect associations in the tails of distributions, and asymmetries of such associations. 
Tail association (introduced formally in the section on Background, below) 
is association between extreme values of variables. 
If smaller values of two positively associated variables are more 
strongly associated than are larger values, the variables are said to
have stronger lower- or left-tail association than right- or upper-tail association
(Fig. \ref{fig_cop_pedag1}C);
and *vice versa* if larger values are more strongly associated than smaller
ones (Fig. \ref{fig_cop_pedag1}D). Datasets of the same 
Spearman or Kendall correlation can have a range of tail association 
(Fig. \ref{SM-fig_cop_pedag}), which can be quantified (see sections on Background and
methods for Q1 below). Although copulas can be used to model the entire
complex dependence structure of data, we have found tail association to be
a useful aspect of the full structure of dependence, so tail association 
is one of the main foci of this paper.

The goal of this paper is to explore the potential for 
applications of copulas in ecology. We provide a basic introduction
to copula concepts in the Background section below, but this is a sketch
only - readers can find abbreviated [@Anderson2018; @Genest2007] 
and comprehensive [@nelsen2006_copula; @joe2014_dependence; @MaiScherer2017]
introductions elsewhere. Instead our goal is to 
help determine to what extent ecological understanding may benefit generally from
using copulas. Copulas have already been used very effectively 
in a few studies to illuminate ecological phenomena 
[@Valpine2014; @Anderson2018; @Popovic2018], but such usage is
still rare. Our results suggest that environmental, 
ecological and evolutionary processes 
may commonly generate and transmit complex dependence structures, including asymmetric tail 
associations, and that greater use of copulas can illuminate the underlying processes.
We believe copula approaches are among the tools all ecologists, including tropical
ecologists, should be considering for analysis of their data in the 21st century.

\begin{figure}[!h]
\begin{center}
\includegraphics[width=7in]{./MasterFigure.pdf}
\caption{Summary and guide for analyses presented in the text. Middle boxes
correspond to ecological datasets for which Q1 was examined (in the 
third through
fifth sections of the text, see also Table \ref{tab_data_info}). Upper boxes 
correspond to potential causes of non-normal (non-N) copula structure that
were examined (in the sixth and seventh sections of the text). Lower boxes 
correspond to potential consequences we examined of non-normal copula structure for
ecological understanding and for applications (in the eighth and ninth
sections of the text). Arrow labels (A-D for analyses pertaining 
to causes, Q2; and X-Z for analyses pertaining to consequences, Q3) 
link to locations in the text.\label{MasterFigure}}
\end{center}
\end{figure}

@@

A second reason we may expect understanding of complex dependence structures
to benefit ecological
research is that prior work has demonstrated, using copulas, complex 
stucture in the spatial dependence of environmental 
variables [@Serinaldi2008; @She2018; @Goswami2018;
@li2013]. One may expect an environmental variable 
measured through time in two locations to show strong tail associations between
the locations if
intense meteorological events are also widespread in their effects, as seems 
frequently to be the case: extreme values are associated with intense events,
so happen in both places at the same time, whereas moderate values of the 
environmental variable may be associated with local phenomena, which differ
between the locations.
Spatial dependence in an environmental variable 
tends to beget spatial dependence 
in fluctuations of populations or other ecosystem variables 
(this is called \emph{spatial synchrony}) which are influenced by the environmental 
variable. This influence is called the \emph{Moran effect}. 
If complex dependence structure is transmitted, 
possibly in modified form, 
from environmental to ecological variables, then we may expect complex dependence 
structure to be a common feature of the 
spatial synchrony of population, community, 
biogeochemical and other environmentally influenced ecological variables.
Synchrony is a phenomenon of substantial interest in 
ecology [@Liebhold2004; @Sheppard2016; @walteretal2017].

A third reason we may expect an understanding of complex 
dependence structures to benefit 
ecological research is that such an approach may help illuminate 
the causal mechanisms between variables. For instance, if two species
Sp1 and Sp2 are strong competitors, abundances of the two species
across quadrats should be negatively related, as in 
Fig. \ref{fig_cop_pedag1}E-F. If Sp1 is plotted on the horizontal axis
and Sp2 on the vertical, then Fig. \ref{fig_cop_pedag1}E 
may suggest Sp1 is the dominant 
competitor: when Sp1 is abundant, Sp2 is necessarily rare because it is suppressed; 
whereas when Sp1 is rare, Sp2
may be abundant, or may also be relatively rare due to 
limiting factors other than Sp1. 
Alternatively, Fig. \ref{fig_cop_pedag1}F may suggest that Sp2
is the dominant competitor. Other causal hypotheses may produce 
similar dependence structure, and it is usually impossible to obtain
complete information on causal pathways from analyses which are
fundamentally based on associations. However, more detailed 
examination of the patterns of dependence of variables may
well be combined with biological and other reasoning to rule out
some causal hypotheses which could not be eliminated solely 
with standard correlation approaches. @Popovic2018 has previously
used copulas to help illuminate causal aspects of relationships
between species.

A fourth reason we may expect a copula understanding of complex 
dependence structures to be useful
for ecology has to do with spatially aggregated or averaged quantities. 
Many ecological variables of applied importance 
depend on the spatial average or sum of local 
quantities. For instance, regional methane and CO$_2$ 
fluxes are the sum of local fluxes; 
and the total economic value of a fishery is the sum of local catches. 
We will explore how details, which can be captured using copulas, 
of the dependence between fluctuations of local
quantities can influence fluctuations of 
the spatial mean or sum, and how this may influence 
higher organizational levels in ecology and human concerns.
Illustrating the idea here, prior work demonstrated that the overall 
reliability of wind-generated electricity depends sensitively on details
of the dependence between wind speeds at multiple generator sites [@li2013]. 
Spatially aggregated ecological variables may be subject to a similar 
effect, with
possible consequences. For instance, if populations of a 
pest species in different locations are all positively associated and are also more strongly
related to each other in their right tails, then local outbreaks will tend 
to occur together, creating regional epidemics. 
On the other hand, stronger left-tail associations in a pest species, even if overall 
correlation were unchanged, would have more benign effects.

We approached our overall goal of exploring the potential of copula approaches 
for ecology by addressing the following specific questions. (Q1) Do datasets
in ecology have dependence structure distinct from that of a multivariate 
Gaussian/normal
distribution (here called non-normal copula structure)? Do positvely associated 
ecological variables show tail 
associations distinct from those of a normal distribution, and in particular
do they show asymmetric tail associations? Normal copula structure is assumed
by standard approaches that use multivariate 
normal distributions or distributions obtained by transforming
the marginals of a normal distribution. So Q1 asks whether ecological
data contain dependency information distinct from a standard or default case. 
(Q2) If the answer to
Q1 is "yes," then what are some possible causes/mechanisms of non-normal
copula structure and asymmetric tail associations in ecology? And (Q3) what are the 
consequences of non-normal copula
structure and asymmetric tail associations for ecological understanding and
applications? After a Background section (the second section), we address Q1 
(third through fifth sections of the paper) 
by analyzing several datasets, including 
environmental, species-trait, phenological, population, community, and 
ecosystem functioning data, selected to span multiple 
sub-fields and organizational levels of ecology. 
We address Q2 (sixth and seventh sections of the paper), using 
simple models. We address Q3 (eighth and ninth sections of the paper) using 
both data and models. The final section is the Discussion.
Multiple analyses are brought to bear on each question, and these
are summarized diagramatically in Fig. \ref{MasterFigure}, which can 
serve as a *post hoc* guide to the paper. 
Copula approaches have been used to great effect in neuroscience [@onken2009], 
bioinformatics [@Kim2008], medical research [@Emura2016], direct study of environmental 
variables [@Serinaldi2008; @She2018; @Goswami2018; @li2013], and finance [@Li2000],
and have also been used very effectively, but rarely so far, in ecology [@Valpine2014; 
@Anderson2018; @Popovic2018]. 
We argue that benefits of wider usage of complete 
copula descriptions of dependence in ecology will be substantial.

# Background on copulas

@@

\begin{figure}[!h]
\begin{center}
\includegraphics[width=6.5in]{./Results/PedagogFigCopThry.pdf}
\caption{Normal (A) and Clayton (B) copulas were combined with standard normal 
marginals (C, D) and gamma marginals (E, F) via Sklar's theorem to produce 
bivariate distributions. Note that a normal copula is distinct from a normal marginal 
distribution: a normal copula is the copula of a bivariate normal distribution.
See the Background section for more information on normal, Clayton,
and other copulas. Each copula was used with both sets of marginals and each
set of marginals was used with both copulas, to demonstrate that: both copula and 
marginals contribute fundamentally to the resulting distribution, the copula 
contributing the information on the dependence between the variables; and that 
one can select the copula and marginals independently. Bivariate
distributions, including the copulas, are depicted via their log-scale probability density functions (pdfs), 
and marginals are depicted via their linear-scale pdfs. Grey dots are
50 random samples from each distribution. The parameter $0.7$ was used for the normal 
copula, the parameter $2$ for the Clayton 
copula, and shape and scale parameters $5$ and $1$ for the gamma marginals. 
Variables $u$ and $v$ were used for copulas
and $x$ and $y$ for distributions created by pairing a copula with
(non-uniform) marginals.
\label{PedagogFigCopThry}}
\end{center}
\end{figure}

We introduce some standard copula families, as examples of copulas and also
because we will fit these families to data later. The bivariate normal
copula family, which consists of copulas of bivariate normal
distributions, is a one-parameter family. The parameter, which we here call $p$,
corresponds to the correlation
of the related bivariate normal distribution, and controls the degree of association
between the variables. A normal copula with $p=0.7$ was already introduced (Fig.
\ref{PedagogFigCopThry}A). The formula for the
copula is $\Phi_{2,p}(\Phi_{1}^{-1}(u),\Phi_{1}^{-1}(v))$, where $\Phi_1^{-1}$ is the inverse of the
cdf of a (univariate) standard normal and $\Phi_{2,p}$ is the joint cdf of a
bivariate normal distribution with mean $(0,0)$ and covariance matrix having $1$s on the diagonal
and $p$ in the off diagonal positions. However, formulas for copulas are often not as 
directly informative about copula properties as pdf diagrams; Fig. \ref{figped_N} has 
example pdfs corresponding to bivariate normal copulas for various values of $p$.
Note that the pdfs are symmetric
around the diagonal line $v=-u+1$ (Fig. \ref{figped_N}), so normal copulas have symmetric 
associations between the two variables in the left and right tails. 
The Clayton copula family,
of which we also already pictured an example (Fig.
\ref{PedagogFigCopThry}B), is another one-parameter family. In contrast to normal copulas, Clayton copulas
have stronger left- than right-tail association. The formula is 
$\left[  \reumax(u^{-p}+v^{-p}-1,0)\right]^{-1/p}$, for parameter $p$; though this again provides 
less direct intuition than do example pdfs (Fig. \ref{figped_C}). Higher values of the parameter,
$p$ produce copulas from the Clayton family with higher Kendall or Spearman correlation. The survival
Clayton family is the symmetric opposite of the Clayton family, showing
stronger right- than left tail association (Fig. \ref{SM-figped_SC}). The BB1 copula 
family is a two-parameter family, thereby providing more flexibility: for some parameters
BB1 copulas have stronger left- than right-tail association, and for others the reverse
(Fig. \ref{figped_BB1}). See @nelsen2006_copula and @joe2014_dependence for further
information on these and other copula families.

>>>>

---
title: "Copulas and their potential for ecology"
author: "Shyamolina Ghosh, Lawrence W. Sheppard, Mark T. Holder, Terrance D. Loecke, Philip C. Reid, James D. Bever, Daniel C. Reuman"
date: ""
fontsize: 12pt
geometry: "left=2.54cm,right=2.54cm,top=2.54cm,bottom=2.54cm"

output: 
  pdf_document:
    number_sections: yes
    keep_tex: yes
    fig_caption: yes
    includes:
      in_header: head_maintext.sty

@@

\noindent \emph{The number of words in the abstract:} 263

@@

<!--***DAN: remianing to do:
DONE 1) Edit abstract
DONE 2) Go through the whole thing one more time
DONE 3) send it out to coauthors (done) and maybe get feedback
DONE 3.5) ESP FROM lAWRENCE
DONE 3.6) Incorporate Shya feedback
DONE 3.7) Incorporate Reid feedback
DONE 4)make sure to say something about tropical systems
DONE 5) maybe write to the editors saying this is coming soon?
DONE 6) In the Discussion, you maybe need to point out not only that we
use only bivariate copulas (and there are multivariate ones), but also
that we use only continuous marginals. Probably an adequate approximation
for our data, but see the other papers I've seen recently which use count data,
and discrete marginals. Approaches such as that will be necessary for some 
datasets.
DONE 7) change the box around the Cedar Creek data in the main fig
DONE 8) say in Discussion that you could have used non-ranks for the nonparam stats - this
is the stuff the Ecol Mongr referee talked about. Maybe this should go in Background instead?
DONE 9) maybe go back and read the Ecol Monogr reviews again
DONE 10) try to emphasize early on that the focus is on positively associated variables, so 'lower tail' means the lower tail of both variables and it is not necessary to worry about negative relationships (in practice we would always transform these for ease).   The dfinitions of tail dependence we use cannot usefully be applied to untransformed data in fig 1e,f.
DONE 11) Number the sections and subsections in the main text and change all the section 
references to use those numbers. Search for "section" (DONE) also for "methods" (DONE)
-->

\noindent \textbf{Abstract}

\noindent All branches of ecology study relationships among and between environmental and biological variables. However, 
standard approaches to studying such relationships, based on correlation and regression, 
provide only a small slice of the complex information contained in the relationships. Other 
statistical approaches exist that provide a complete description of relationships between variables, 
based on the concept of the *copula*; they are applied in finance, neuroscience and other fields, 
but rarely in ecology. We here explore the concepts that underpin copulas and examine the potential 
for those concepts to improve our understanding of ecology. We find that informative copula 
structure in dependencies between variables is common across all the environmental, 
species-trait, phenological, population, community, and ecosystem functioning datasets we 
considered. Many datasets exhibited asymmetric tail associations, whereby two variables were more 
strongly related in their left compared to right tails, or *vice versa*. We describe mechanisms by which 
observed copula structure and asymmetric tail associations can arise in ecological data, including a Moran-like 
effect whereby dependence structures between environmental variables are inherited by ecological 
variables; and asymmetric or nonlinear influences of environments on ecological variables, such as 
under Liebig's law of the minimum. We also describe consequences of copula structure for 
ecological phenomena, including impacts 
on extinction risk, Taylor's law, and the stability through time of ecosystem services. By documenting the 
importance of a complete description of dependence between variables, advancing conceptual frameworks, and 
demonstrating a powerful approach, we aim to encourage widespread use of copulas in ecology, which we
believe can benefit the discipline.

@@

All branches of ecology 
study relationships among biological variables and relationships between 
environmental and biological 
variables. However, commonly used correlation and regression 
approaches to studying 
such relationships are limited, and 
provide only a partial description of a relationship between variables. 
For instance, datasets showing markedly different 
relationships may have the same correlation coefficient
(Fig. \ref{fig_cop_pedag1}A-B). The variables of Fig. \ref{fig_cop_pedag1}A 
(respectively, Fig. \ref{fig_cop_pedag1}B) are principally 
related in the left (respectively, right) portions of their distributions, 
an asymmetric pattern of dependence that can
have ecological significance, as discussed below, but that is not captured by 
correlation. Although correlation is only one way to 
study relationships between variables, other common metrics also 
provide only partial information.

@@

Well-developed approaches do exist, however, and are 
applied widely in other fields, 
that provide a description of the dependence 
between two or more variables that can be mathematically proven 
to be complete [@nelsen2006_copula; @joe2014_dependence; @MaiScherer2017]. 
The approaches are based on 
the concept of the *copula*. Copula approaches separate the information of a bivariate 
random variable into two distinct parts: the information in the 
marginal distributions (which says nothing about the dependence between the 
variables), and the remaining information (which is solely about
dependence). We here introduce some simple concepts, based on taking ranks,
that relate to copulas and give a conceptual 
flavor of copula ideas to be introduced 
more formally in the "Background on copulas" section below (section \ref{Background}). Given 
a sample $(x_i,y_i)$, $i=1,2,\ldots,n$ (e.g., 
Fig. \ref{fig_cop_pedag1}A-B), information about the structure of the dependence
between $x$ and $y$ can be separated from information contained 
in the marginal distributions by 
considering the bivariate plot of $u_i$ versus $v_i$, where
$u_i$ is the rank of $x_i$ in the set $x_j$ ($j=1,\ldots,n$), divided by
$n+1$; and $v_i$ is defined in the same way but using the $y_i$. The rank
of the smallest element in a set is $1$. The $u_i$ and 
$v_i$ are called *normalized ranks* of the $x_i$ and $y_i$; they relate to 
the empirical cumulative distribution function of the $x_i$ and $y_i$,
respectively. We henceforth abbreviate "cumulative distribution function" as "distribution
function" or "cdf". See, for instance,
Fig. \ref{fig_cop_pedag1}C-D, which show the normalized ranks of 
A-B, respectively. Ranking makes the marginal distributions of 
the component datasets uniform, isolating the 
dependence structure. Dependence structure and marginals can then be studied 
separately. Normalized rank plots such as Fig. \ref{fig_cop_pedag1}C-D
relate to copulas in a way to be described in the "Background on copulas" section below
(section \ref{Background}), after a definition of copulas
is provided. Note that Pearson correlation, even though it is the most commonly
used measure of relationships between variables, is modified by monotonic
tranformation of the component variables (Appendix 
\ref{SM-pedagog1_details}, Fig. \ref{SM-fig_transf_pearson_spearman}),
and therefore reflects not only dependence information, but also
information on the marginals [@Genest2007]. 
Spearman and Kendall correlations, being rank-based, 
are not influenced by monotonic transformations of the component variables. 
They provide information solely on the dependence between those variables.

A main benefit of a copula approach is 
that it can detect associations in the tails of distributions, and asymmetries of such associations. 
Tail association (introduced formally in the "Background on copulas" section, below, section 
\ref{Background}) 
is association between extreme values of variables. 
If smaller values of two positively associated variables are more 
strongly associated than are larger values, the variables are said to
have stronger lower- or left-tail association than right- or upper-tail association
(Fig. \ref{fig_cop_pedag1}C);
and *vice versa* if larger values are more strongly associated than smaller
ones (Fig. \ref{fig_cop_pedag1}D). Datasets of the same 
Spearman or Kendall correlation can have a range of tail associations 
(Fig. \ref{SM-fig_cop_pedag}), which can be quantified (see sections on Background and
methods for Q1 below, sections \ref{Background} and \ref{Methods}). 
Although copulas can be used to model the entire
complex dependence structure of data, we have found tail associations to
usefully describe the full structure of dependence, so tail association 
is one of the main foci of this paper.

The goal of this paper is to explore the potential for 
applications of copulas in ecology and to 
help determine to what extent ecological understanding may benefit generally from
using copulas. We provide a basic introduction
to copula concepts in the "Background on copulas" section below (section \ref{Background}). 
Readers can find additional abbreviated [@Anderson2018; @Genest2007] 
and comprehensive [@nelsen2006_copula; @joe2014_dependence; @MaiScherer2017]
introductions elsewhere. Copulas have already been used very effectively 
in a few studies to illuminate ecological phenomena 
[@Valpine2014; @Anderson2018; @Popovic2018], but such usage is
still rare. Our results suggest that environmental, 
ecological and evolutionary processes 
may commonly generate and transmit complex dependence structures, including asymmetric tail 
associations, and that greater use of copulas can illuminate the underlying processes.
We believe copula approaches are among the tools all ecologists, including tropical
ecologists, should be considering for analysis of their data in the 21st century.

\begin{figure}[!h]
\begin{center}
\includegraphics[width=7in]{./MasterFigure.pdf}
\caption{Summary and guide for analyses presented in the text. Middle boxes
correspond to ecological datasets for which Q1 was examined (in sections
\ref{Data}-\ref{Results}, see also Table \ref{tab_data_info}). Upper boxes 
correspond to potential causes of non-normal (non-N) copula structure that
were examined (in sections \ref{Causes}-\ref{Results_Q2} of the text). Lower boxes 
correspond to potential consequences we examined of non-normal copula structure for
ecological understanding and for applications (in sections 
\ref{Consequences}-\ref{Results_Q3} of the text). Arrow labels (A-D for analyses pertaining 
to causes, Q2; and X-Z for analyses pertaining to consequences, Q3) 
link to locations in the text.\label{MasterFigure}}
\end{center}
\end{figure}

@@

A second reason why we may expect understanding of complex dependence structures
to benefit ecological
research is that prior work has demonstrated, using copulas, complex 
stucture in the spatial dependence of environmental 
variables [@Serinaldi2008; @She2018; @Goswami2018;
@li2013]. One may expect an environmental variable 
measured through time in two locations to show strong tail associations between
the locations if
intense meteorological events are also widespread in their effects, as seems 
frequently to be the case: extreme values are associated with intense events,
so happen in both places at the same time, whereas moderate values of the 
environmental variable may be associated with local phenomena, which differ
between the locations.
Spatial dependence in an environmental variable 
tends to beget spatial dependence 
in fluctuations of populations or other ecosystem variables 
(this is called \emph{spatial synchrony}) which are influenced by the environmental 
variable. This influence is called the \emph{Moran effect}. 
If complex dependence structure is transmitted, 
possibly in modified form, 
from environmental to ecological variables, then we may expect complex dependence 
structure to be a common feature of the 
spatial synchrony of population, community, 
biogeochemical and other environmentally influenced ecological variables.
Synchrony is a phenomenon of substantial interest in 
ecology [@Liebhold2004; @Sheppard2016; @walteretal2017].

A third reason why we may expect an understanding of complex 
dependence structures to benefit 
ecological research is that such an approach may help illuminate 
the causal mechanisms between variables. For instance, if two species
Sp1 and Sp2 are strong competitors, abundances of the two species
across quadrats should be negatively related, as in 
Fig. \ref{fig_cop_pedag1}E-F. If Sp1 is plotted on the horizontal axis
and Sp2 on the vertical, then Fig. \ref{fig_cop_pedag1}E 
may suggest Sp1 is the dominant 
competitor: when Sp1 is abundant, Sp2 is necessarily rare because it is suppressed; 
whereas when Sp1 is rare, Sp2
may be abundant, or may also be relatively rare due to 
limiting factors other than Sp1. 
Alternatively, Fig. \ref{fig_cop_pedag1}F may suggest that Sp2
is the dominant competitor. Other causal hypotheses may produce 
similar dependence structure, and it is usually impossible to obtain
complete information on causal pathways from analyses which are
fundamentally based on associations. However, more detailed 
examination of the patterns of dependence of variables may
well be combined with biological and other reasoning to rule out
some causal hypotheses which could not be eliminated solely 
with standard correlation approaches. @Popovic2018 has previously
used copulas to help illuminate causal aspects of relationships
between species.

A fourth reason why we may expect a copula understanding of complex 
dependence structures to be useful
for ecology has to do with spatially aggregated or averaged quantities. 
Many ecological variables of applied importance 
depend on the spatial average or sum of local 
quantities. For instance, regional methane and CO$_2$ 
fluxes are the sum of local fluxes; 
and the total economic value of a fishery is the sum of local catches. 
We will explore how details, which can be captured using copulas, 
of the dependence between fluctuations of local
quantities can influence fluctuations of 
the spatial mean or sum, and how this may influence 
higher organizational levels in ecology and human concerns.
To illustrate this idea we cite @li2013, who have demonstrated that the overall 
reliability of wind-generated electricity depends sensitively on details
of the dependence between wind speeds at multiple generator sites. 
Spatially aggregated ecological variables may be subject to a similar 
effect, with
possible consequences. For instance, if populations of a 
pest species in different locations are all positively associated and are also more strongly
related to each other in their right tails, then local outbreaks will tend 
to occur together, creating regional epidemics. 
On the other hand, stronger left-tail associations in a pest species, even if overall 
correlation were unchanged, would have more benign effects.

We approached our overall goal of exploring the potential of copula approaches 
for ecology by addressing the following specific questions. (Q1) Do datasets
in ecology have dependence structure distinct from that of a multivariate 
Gaussian/normal
distribution (here called non-normal copula structure)? Do positvely associated 
ecological variables show tail 
associations distinct from those of a normal distribution, and in particular
do they show asymmetric tail associations? Normal copula structure is assumed
by standard approaches that use multivariate 
normal distributions or distributions obtained by transforming
the marginals of a normal distribution. So Q1 asks whether ecological
data contain dependency information distinct from a standard or default case. 
(Q2) If the answer to
Q1 is "yes," then what are some possible causes/mechanisms of non-normal
copula structure and asymmetric tail associations in ecology? And (Q3), what are the 
consequences of non-normal copula
structure and asymmetric tail associations for ecological understanding and
applications? After section \ref{Background}, 'Background on copulas', we address Q1 
(sections \ref{Data} through \ref{Results} of the paper) 
by analyzing several datasets, including 
environmental, species-trait, phenological, population, community, and 
ecosystem functioning data, selected to span multiple 
sub-fields and organizational levels of ecology. 
We address Q2 (sections \ref{Causes} through \ref{Results_Q2} of the paper), using 
simple models. We address Q3 (sections \ref{Consequences} through \ref{Results_Q3} 
of the paper) using 
both data and models. Section \ref{Discussion} is the Discussion.
Multiple analyses are brought to bear on each question, and these
are summarized diagramatically in Fig. \ref{MasterFigure}, which can 
serve as a *post hoc* guide to the paper. 
Copula approaches have been used to great effect in neuroscience [@onken2009], 
bioinformatics [@Kim2008], medical research [@Emura2016], direct study of environmental 
variables [@Serinaldi2008; @She2018; @Goswami2018; @li2013], and finance [@Li2000],
and have also been used very effectively, but rarely so far, in ecology [@Valpine2014; 
@Anderson2018; @Popovic2018]. 
We argue that benefits of wider usage of complete 
copula descriptions of dependence in ecology will be substantial.

# Background on copulas\label{Background}

@@

\begin{figure}[!h]
\begin{center}
\includegraphics[width=6.5in]{./Results/PedagogFigCopThry.pdf}
\caption{Normal (A) and Clayton (B) copulas were combined with standard normal 
marginals (C, D) and gamma marginals (E, F) via Sklar's theorem to produce 
bivariate distributions. Note that a normal copula is distinct from a normal marginal 
distribution: a normal copula is the copula of a bivariate normal distribution.
See section \ref{Background} for more information on normal, Clayton,
and other copulas. Each copula was used with both sets of marginals and each
set of marginals was used with both copulas, to demonstrate that: both copula and 
marginals contribute fundamentally to the resulting distribution, the copula 
contributing the information on the dependence between the variables; and that 
one can select the copula and marginals independently. Bivariate
distributions, including the copulas, are depicted via their log-scale probability density functions (pdfs), 
and marginals are depicted via their linear-scale pdfs. Grey dots are
50 random samples from each distribution. The parameter $0.7$ was used for the normal 
copula, the parameter $2$ for the Clayton 
copula, and shape and scale parameters $5$ and $1$ for the gamma marginals. 
Variables $u$ and $v$ were used for copulas
and $x$ and $y$ for distributions created by pairing a copula with
(non-uniform) marginals.
\label{PedagogFigCopThry}}
\end{center}
\end{figure}

We introduce some standard copula families, as examples of copulas and also
because we will fit these families to data later. The bivariate normal
copula family, which consists of copulas of bivariate normal
distributions, is a one-parameter family. The parameter, which we here call $p$,
corresponds to the correlation
of the related bivariate normal distribution, and controls the degree of association
between the variables. A normal copula with $p=0.7$ was already introduced (Fig.
\ref{PedagogFigCopThry}A). The formula for the
copula is $\Phi_{2,p}(\Phi_{1}^{-1}(u),\Phi_{1}^{-1}(v))$, where $\Phi_1^{-1}$ is the inverse of the
cummulative dstribution function (cdf) of a (univariate) standard 
normal and $\Phi_{2,p}$ is the joint cdf of a
bivariate normal distribution with mean $(0,0)$ and covariance matrix having $1$s on the diagonal
and $p$ in the off diagonal positions. However, formulas for copulas are often not as 
directly informative about copula properties as probability density function (pdf) 
pictures; Fig. \ref{figped_N} has 
example pdfs corresponding to bivariate normal copulas for various values of $p$.
Note that the pdfs are symmetric
around the diagonal line $v=-u+1$ (Fig. \ref{figped_N}), so normal copulas have symmetric 
associations between the two variables in the left and right tails. 
The Clayton copula family,
of which we have already pictured an example (Fig.
\ref{PedagogFigCopThry}B), is another one-parameter family. In contrast to normal copulas, Clayton copulas
have stronger left- than right-tail association. The formula is 
$\left[  \reumax(u^{-p}+v^{-p}-1,0)\right]^{-1/p}$, for parameter $p$; though this again provides 
less direct intuition than do example pdfs (Fig. \ref{figped_C}). Higher values of the parameter,
$p$ produce copulas from the Clayton family with higher Kendall or Spearman correlation. The survival
Clayton family is the symmetric opposite of the Clayton family, showing
stronger right- than left tail association (Fig. \ref{SM-figped_SC}). The BB1 copula 
family is a two-parameter family, thereby providing more flexibility: for some parameters
BB1 copulas have stronger left- than right-tail association, and for others the reverse
(Fig. \ref{figped_BB1}). See @nelsen2006_copula and @joe2014_dependence for further
information on these and other copula families.
